\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-12 jasa/paper-jasa.tex

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{theorem}
\usepackage{xfrac}

@@

The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters. The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$.
Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical.
Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\cite{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available. In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.

As a concrete example, consider searches for new particles at the Large Hadron Collider.
The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\cite{Sjostrand:2006za,Agostinelli:2002hh}.

The ATLAS and CMS experiments have published  hundreds papers where the
final  result was formulated as a hypothesis test or confidence interval
using a generalized  likelihood ratio test~\cite{Cowan:2010js}. This includes
the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and
subsequent measurement of its properties.

The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes
(labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\cite{Cranmer:2012sba}.

To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history.

While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g. the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.

The initial motivation for this work was to extend the typical usage of discriminative classifiers in HEP to be robust to
nuisance parameters in the simulators. The scope of the result expanded once it became clear that this offers a way to approximate the likelihood function $p(x|\theta)$ in what is typically considered the likelihood-free setting. This approach is complementary to ABC as it does not require a prior over the parameters and can also be used in the classical (frequentist) setting. A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.

@@

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
\begin{equation}
T(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case
high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott}, \cite{JMLR:v14:tong13a}, and \cite{Neal:2007zz}.

@@

Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}. Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative classifiers model the posterior $p(y|x)$ directly. For classification tasks, one then thresholds on $p(y|x)$. In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data.
This familiar formulation of classification may lead to some confusion in the setting of the current work.

@@

Lastly, in the setting of frequentist hypothesis tests and confidence intervals, we do not have a prior $\pi(\theta)$.
While we can use the generative models to produce training data $(x,y=0)$ generated
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.

@@

One strength of this approach is that it factorizes the approximation of the per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor job of reproducing the level sets of the per-event likelihood ratio, the density of $\hat{s}$ can still be well calibrated. In that case, one might loose power, but the resulting inference will still be valid. This point was made by \cite{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.

@@

Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
For an single parameter point, this is a tractable univariate density estimation problem. The challenge comes from the need to estimate this density for all values of $\theta$.
A straight forward approach would be to run the generative model on demand for any particular value of $\theta$.
In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$. This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).  In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\cite{Cranmer:2012sba}.

There is a mild technical challenge in embedding the classifier into the likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$. Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.  A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

In HEP we are often searching for some
class of events, generically referred to as \textit{signal}, in the presence of a separate class
of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions
$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
uncertainties in the underlying physics prediction or response of the measurement device. The
total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate
to the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}. This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\cite{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}. In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time to calculate the likelihood ratio $q(x)$ for a single event. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.  Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Gao:2010qx,Bolognesi:2012mm}.

@@

\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing. As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d. samples from each hypothesis. This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification. Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
In a similarly titled work, \cite{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC. While the goal there is also parameter inference in the likelihood-free setting,  ``classifier ABC'' is very different than the approach presented here.
\cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods. This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
%\cite{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
\cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this. In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests. \cite{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.

\cite{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available. That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
The key distinction comes in the handling of $\theta$.  Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.  The work of Neal correctly identifies this as an approximation of the target likelihood even in the case of a ideal classifier. In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
\footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is explicitly being tested. This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier. In both cases, the quality of the classifier is factorized from the calibration of its density, which allows for valid inference even if there is a loss of power due to a non ideal classifier.

@@

\section{References}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{theorem}
\usepackage{xfrac}
\usepackage{hyperref}

@@

The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters. The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$.
Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical.
Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citep{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available. In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.

As a concrete example, consider searches for new particles at the Large Hadron Collider.
The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}.

The ATLAS and CMS experiments have published  hundreds papers where the
final  result was formulated as a hypothesis test or confidence interval
using a generalized  likelihood ratio test~\citep{Cowan:2010js}. This includes
the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and
subsequent measurement of its properties.

The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes
(labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\citep{Cranmer:2012sba}.

To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.

While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g. the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.

The initial motivation for this work was to extend the typical usage of discriminative classifiers in HEP to be robust to
nuisance parameters in the simulators. The scope of the result expanded once it became clear that this offers a way to approximate the likelihood function $p(x|\theta)$ in what is typically considered the likelihood-free setting. This approach is complementary to ABC as it does not require a prior over the parameters and can also be used in the classical (frequentist) setting. A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme sketched by \citep{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.

@@

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
\begin{equation}
T(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case
high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \citep{ClaytonScott}, \citep{JMLR:v14:tong13a}, and \citep{Neal:2007zz}.

@@

Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative classifiers model the posterior $p(y|x)$ directly. For classification tasks, one then thresholds on $p(y|x)$. In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data.
This familiar formulation of classification may lead to some confusion in the setting of the current work.

@@

Lastly, in the setting of frequentist hypothesis tests and confidence intervals, we do not have a prior $\pi(\theta)$.
While we can use the generative models to produce training data $(x,y=0)$ generated
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.  Similar points have been made by~\citep{ClaytonScott} and \citep{Neal:2007zz}.

@@

One strength of this approach is that it factorizes the approximation of the per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor job of reproducing the level sets of the per-event likelihood ratio, the density of $\hat{s}$ can still be well calibrated. In that case, one might loose power, but the resulting inference will still be valid. This point was made by \citep{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.

@@

Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
For an single parameter point, this is a tractable univariate density estimation problem. The challenge comes from the need to estimate this density for all values of $\theta$.
A straight forward approach would be to run the generative model on demand for any particular value of $\theta$.
In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$. This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).  In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.

There is a mild technical challenge in embedding the classifier into the likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$. Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.  A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

In HEP we are often searching for some
class of events, generically referred to as \textit{signal}, in the presence of a separate class
of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions
$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
uncertainties in the underlying physics prediction or response of the measurement device. The
total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate
to the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}. In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time to calculate the likelihood ratio $q(x)$ for a single event. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.  Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.

@@

\citep{ClaytonScott} and \citep{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing. As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d. samples from each hypothesis. This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification. Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
In a similarly titled work, \citep{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC. While the goal there is also parameter inference in the likelihood-free setting,  ``classifier ABC'' is very different than the approach presented here.
\citep{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods. This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
%\citep{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
\citep{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this. In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests. \citep{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.

\citep{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available. That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
The key distinction comes in the handling of $\theta$.  Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.  The work of Neal correctly identifies this as an approximation of the target likelihood even in the case of a ideal classifier. In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
\footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is explicitly being tested. This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier. In both cases, the quality of the classifier is factorized from the calibration of its density, which allows for valid inference even if there is a loss of power due to a non ideal classifier.

@@
